---
title: "Lab 5 - Monte Carlo Feature Selection"
subtitle: "Knowledge-based Systems"
author: "Rasmus Hammar"
format: typst
---

## Library

```{r}
library(rmcfs)
```

## Data

```{r}
# load dataset
alizadeh <- read.table("alizadeh.csv")
```

```{r}
#| tbl-cap: Preview of alizadeh data.
#| label: tab-data_preview
n = 5
alizadeh[1:n, c(1:n, dim(alizadeh)[2])] |>
  knitr::kable()
```

The universe is all sets/rows in the dataframe.

The attributes are all of the 'GENE' columns, and the decision attribute is the 'class' column.

```{r}
table(alizadeh$class) |>
  as.data.frame() |>
  `names<-`(c("class", "Frequency")) |>
  knitr::kable()
```

## Monte Carlo Feature Selection


```{r}
result <- mcfs(
  class ~ .,
  alizadeh,
  projections = 1500,
  projectionSize = 0.1,
  splits = 5,
  splitSetSize = 500,
  cutoffPermutations = 6,
  threadsNumber = 6
)
```

* projections = 1500 => number of random projections to be generated
* projectionSize = 0.1 => size of the random projections, i.e. how many columns (10 %)
* splits = 5 => number of splits, dividing rows into chunks, for every projection
* splitSetSize = 500 => size of the splits, i.e. how many rows
* cutoffPermutations = 6 => number of permutations to calculate the cutoff

```{r}
head(result$RI) |>
  knitr::kable(row.names = F)
```

As seen in @fig-mcfs_convergence, the rankings stabalize after around 1100 projections. Re-running the algorithm would likely not change the results much.

```{r}
#| fig-cap: MCFS convergence over 1500 permutations.
#| label: fig-mcfs_convergence
plot(result, type = "distances")
```

Classification matrix with rows as true and columns as predicted classes.

Accuracy = sum(matrix diagonal) / sum(matrix)

True positives = class / sum(row)

False positives = class / sum(column)

The top 30 results are mostly the same as in the paper (@tbl-ri_result_sorted).

```{r}
#| tbl-cap: Top 30 results.
#| label: tbl-ri_result_sorted
result2 <- result$RI[1:result$cutoff_value, ]
result2 |>
  dplyr::arrange(desc(RI)) |>
  dplyr::filter(RI > 0.05) |>
  head(30) |>
  knitr::kable()
```

Difference between the lowest significant and highest non-significant gene is 1.420021 (@tbl-low_high).

```{r}
#| tbl-cap: Lowest significant and highest non-significant genes.
#| label: tbl-low_high
signif_low <- result2 |>
  dplyr::arrange(RI) |>
  dplyr::filter(RI > 0.05)
nonSignif_high <- result$RI |>
  dplyr::arrange(desc(RI)) |>
  dplyr::filter(RI <= 0.05)
low_high <- rbind(signif_low[1, ], nonSignif_high[1, ])
low_high |>
  knitr::kable()
```

The interdependency graph (@fig-interdependencies) shows the top 20 genes and how they depend on each other.

```{r}
#| fig-cap: Interdependency graph.
#| label: fig-interdependencies
gid <- build.idgraph(result, size = 20)
plot.idgraph(gid, label_dist = 0.3)
```
