---
title: "Lab 3"
subtitle: "Knowledge-Based Systems in Bioinformatics"
author: "Rasmus Hammar"
format: typst
---

## Tasks: Discretization

A = (U,A ∪ {d})

```{r}
info_TA <- c(
  -(4 / 9) * log2(4 / 9) - (5 / 9) * log2(5 / 9)
)
```

Info([4,5]) = `r info_TA` bits.

Table 1. Decision system A discretized using equal frequency with three levels.

|index| Cuts feature_1 | decision  |
|-----|-----------|-----------|
|x~6~| [91, 105) | No        |
|x~7~| [91, 105) | No        |
|x~1~| [91, 105) | No        |
|x~9~| [45, 91)  | No        |
|x~8~| [45, 91)  | Yes       |
|x~2~| [45, 91)  | No        |
|x~4~| [21, 45)  | Yes       |
|x~3~| [21, 45)  | Yes       |
|x~5~| [21, 45)  | Yes       |

POS~feature_1~({d}) = {{x~1~, x~6~, x~7~}, {x~3~, x~4~, x~5~}} $=> \gamma = \frac{6}{9}$

Table 2. Decision system A discretized using naive method.

|index| Cuts feature_1 | decision |
|-----|-----------|----------|
|x~6~| [77.5, 105) | No       |
|x~7~| [77.5, 105) | No       |
|x~1~| [77.5, 105) | No       |
|x~9~| [77.5, 105) | No       |
|x~8~| [61.5, 77.5)  | Yes      |
|x~2~| [45, 61.5)  | No       |
|x~4~| [21, 45)  | Yes      |
|x~3~| [21, 45)  | Yes      |
|x~5~| [21, 45)  | Yes      |

POS~feature_1~({d}) = {{x~1~, x~6~, x~7~, x~9~}, {x~8~}, {x~2~}, {x~3~, x~4~, x~5~}} $=> \gamma = \frac{9}{9} = 1$

Comparing $\gamma = \frac{6}{9}$ and $\gamma = \frac{9}{9} = 1$, the naive method would be better, but the equal frequency method is probably good enough and more generalizable.

B = (U,B ∪ {d})

Table 3. Decision system A discretized using equal frequency with three levels.

|index| Cuts width [mm] | decision |
|-----|--------------|----------|
|x~8~| [0.67, 0.97) | A   |
|x~1~| [0.67, 0.97) | B   |
|x~4~| [0.67, 0.97) | A   |
|x~3~| [0.29, 0.67) | A   |
|x~2~| [0.29, 0.67) | A   |
|x~9~| [0.29, 0.67) | B   |
|x~7~| [0.04, 0.29) | B   |
|x~6~| [0.04, 0.29) | B   |
|x~5~| [0.04, 0.29) | A   |
|x~10~| [0.04, 0.29) | B   |

POS~width~({d}) = $\emptyset => \gamma = 0$

Table 4. Decision system A discretized using naive method.

|index| Cuts width [mm] | decision  |
|-----|--------------|-----------|
|x~8~| [0.89, 0.97) | A         |
|x~1~| [0.75, 0.89) | B         |
|x~4~| [0.39, 0.75) | A         |
|x~3~| [0.39, 0.75) | A         |
|x~2~| [0.39, 0.75) | A         |
|x~9~| [0.17, 0.39) | B         |
|x~7~| [0.17, 0.39) | B         |
|x~6~| [0.17, 0.39) | B         |
|x~5~| [0.08, 0.17) | A         |
|x~10~| [0.04, 0.08) | B         |

POS~width~({d}) = {{x~8~}, {x~1~}, {x~4~, x~3~, x~2~}, {x~9~, x~7~, x~6~}, {x~5~}, {x~10~}} => $\gamma = 1$

Comparing $\gamma = 0$ and $\gamma = 1$, the naive method would be better. Equal frequency method is probably good enough and more generalizable, and could probably be reduced to [0.04, 0.29) and [0.29, 0.97).


## Tasks: R.ROSETTA

```{r}
rlang::is_installed("R.ROSETTA")
```

```{r}
library(R.ROSETTA)
```

a)
n features = 35

b)
n objects per class = 146
